---
title: 'GOF'
author: "Matteo Calgaro"
date: 'Compiled: `r format(Sys.Date(), "%d %B, %Y")`'
output:
  html_document:
    code_folding: hide
    df_print: kable
    number_sections: no
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

<style>
#TOC {
  top: 1%;
  opacity: 0.5;
}
#TOC:hover {
  opacity: 1;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#source("additional_functions.R")
```

# Data loading

Data from HMP16SData and curatedMetagenomicData

```{r}
library(HMP16SData)
library(curatedMetagenomicData)
library(phyloseq)
library(reshape2)
library(ggplot2)
library(cowplot)
```

## HMP16SData

Keep features with more than 10 counts in more than 1 samples.
A dataset is composed by samples from a HMP_BODY_SUBSITE.

```{r}
if(!file.exists("../data/HMP16SData_list.RData")){
  V35_all <- V35() %>% as_phyloseq()
  subsites <- unique(V35_all@sam_data$HMP_BODY_SUBSITE)
  HMP16SData_list <- list()
  for(i in 1:length(subsites)){
    cat("Subsite:",subsites[i],"\n")
    # Keep bodysite
    ps <- subset_samples(V35_all,HMP_BODY_SUBSITE == subsites[i])
    # Keep one sample for each Random Subject IDentifier
    ps <- subset_samples(ps,!duplicated(RSID))
    ps <- prune_samples(sample_sums(ps) >= 10^3, ps)
    ps <- filter_taxa(ps,function(x) sum(x>10)>1,1)
    HMP16SData_list[[i]] <- ps
    names(HMP16SData_list)[i] <- subsites[i]
  }
  save(HMP16SData_list,file = "../data/HMP16SData_list.RData")
} else load("../data/HMP16SData_list.RData")
names(HMP16SData_list)
```

## HMPWMSData

Keep features with more than 10 counts in more than 1 samples.
A dataset is composed by samples from a body_subsite.

```{r}
if(!file.exists("../data/HMPWMSData_list.RData")){
  HMPWMSData_list <- list()
  HMPWMSData <- curatedMetagenomicData("HMP_2012.metaphlan_bugs_list*",
                       dryrun = FALSE,
                       counts = TRUE,
                       bugs.as.phyloseq = TRUE)
  for(i in 1:length(HMPWMSData@listData)){
    body_subsites <- unique(HMPWMSData@listData[[i]]@sam_data$body_subsite)
    for(j in 1:length(body_subsites)){
      ps <- HMPWMSData@listData[[i]]
      # Keep body subsite
      ps <- subset_samples(ps,body_subsite == body_subsites[j])
      # Keep 1 sample for each subject
      ps <- subset_samples(ps,!duplicated(subjectID))
      ps <- prune_samples(sample_sums(ps) >= 10^3, ps)
      if(nsamples(ps)>4){
        ps <- filter_taxa(ps,function(x) sum(x>10)>1,1)
        cat(body_subsites[j],ntaxa(ps),"x",nsamples(ps),"\n")
        HMPWMSData_list <- c(HMPWMSData_list,ps)
      }
    }
  }
  names(HMPWMSData_list) <- c("Anterior Nares","Tongue Dorsum","Buccal Mucosa","Supragingival Plaque","Subgingival Plaque","Left Retroauricolar Crease","Right Retroauricolar Crease","Stool","Posterior Fornix")
  save(HMPWMSData_list,file = "../data/HMPWMSData_list.RData")
} else load("../data/HMPWMSData_list.RData")
names(HMPWMSData_list)
```

## AllWMSData

Keep features with more than 10 counts in more than 1 samples.
A dataset is composed by samples from a body site and a single study_condition.

```{r}
if(!file.exists("../data/AllWMSData_list.RData")){
  AllWMSData_list <- list()
  names <- c()
  AllWMSData <- curatedMetagenomicData("*metaphlan_bugs_list*",
                       dryrun = FALSE,
                       counts = TRUE,
                       bugs.as.phyloseq = TRUE)
  
  for(i in 1:length(AllWMSData@listData)){
    body_sites <- unique(AllWMSData@listData[[i]]@sam_data$body_site)
    for(j in 1:length(body_sites)){
      study_conditions <- unique(unlist(AllWMSData@listData[[i]]@sam_data[AllWMSData@listData[[i]]@sam_data$body_site == body_sites[j],"study_condition"]@.Data))
      study_conditions <- study_conditions[!is.na(study_conditions)]
        for(k in 1:length(study_conditions)){
          ps <- AllWMSData@listData[[i]]
          # Keep body subsite
          ps <- subset_samples(ps,body_site == body_sites[j] & study_condition == study_conditions[k])
          # Keep 1 sample for each subject
          ps <- subset_samples(ps,!duplicated(subjectID))
          ps <- prune_samples(sample_sums(ps) >= 10^3, ps)
          if(nsamples(ps)>4){
            ps <- filter_taxa(ps,function(x) sum(x>10)>1,1)
            ps <- prune_samples(sample_sums(ps) > 0, ps)
            name <- paste0(gsub(names(AllWMSData@listData)[[i]], pattern = ".metaphlan_bugs_list.",replacement = "_"),"_",study_conditions[k])
            cat(name,ntaxa(ps),"x",nsamples(ps),"\n")
            ps@sam_data$name <- name
            AllWMSData_list <- c(AllWMSData_list,ps)
            names <- c(names,name)
        }
      }
    }
  }
  names(AllWMSData_list) <- names
  # Dataset used in the paper
  names_paper <- c("AsnicarF_2017_milk_control","AsnicarF_2017_stool_control","Bengtsson-PalmeJ_2015_stool_control","BritoIL_2016_oralcavity_control","BritoIL_2016_stool_control","Castro-NallarE_2015_oralcavity_schizophrenia","Castro-NallarE_2015_oralcavity_control","ChngKR_2016_skin_AD","ChngKR_2016_skin_control","DavidLA_2015_stool_control","DavidLA_2015_stool_infectiousgastroenteritis","FengQ_2015_stool_CRC","FengQ_2015_stool_control","FengQ_2015_stool_adenoma","HanniganGD_2017_stool_adenoma","HanniganGD_2017_stool_control","HanniganGD_2017_stool_CRC","Heitz-BuschartA_2016_stool_T1D","Heitz-BuschartA_2016_stool_control","HMP_2012_nasalcavity_control","HMP_2012_oralcavity_control","HMP_2012_skin_control","HMP_2012_stool_control","HMP_2012_vagina_control","KarlssonFH_2013_stool_IGT","KarlssonFH_2013_stool_control","KarlssonFH_2013_stool_T2D","KosticAD_2015_stool_control","LeChatelierE_2013_stool_control","LiJ_2014_stool_control","LiJ_2017_stool_control","LiJ_2017_stool_pre-hypertension","LiJ_2017_stool_hypertension","LiSS_2016_stool_control","LiSS_2016_stool_metabolic_syndrome","LiSS_2016_stool_FMT","LiuW_2016_stool_control","LomanNJ_2013_stool_STEC","LoombaR_2017_stool_fatty_liver","LouisS_2016_stool_control","NielsenHB_2014_stool_control","NielsenHB_2014_stool_IBD","Obregon-TitoAJ_2015_stool_control","OhJ_2014_skin_control","PasolliE_2018_stool_control","QinJ_2012_stool_control","QinJ_2012_stool_T2D","QinN_2014_stool_control","QinN_2014_stool_cirrhosis","RampelliS_2015_stool_control","RaymondF_2016_stool_control","RaymondF_2016_stool_cephalosporins","SchirmerM_2016_stool_control","ShiB_2015_oralcavity_periodontitis","ShiB_2015_oralcavity_SRP","SmitsSA_2017_stool_control","TettAJ_2016_skin_control","TettAJ_2016_skin_psoriasis","ThomasAM_2018a_stool_control","ThomasAM_2018a_stool_adenoma","ThomasAM_2018a_stool_CRC","VatanenT_2016_stool_control","VatanenT_2016_stool_bronchitis","VatanenT_2016_stool_otitis","VatanenT_2016_stool_respiratoryinf","VincentC_2016_stool_control","VincentC_2016_stool_CDI","VogtmannE_2016_stool_control","VogtmannE_2016_stool_CRC","WenC_2017_stool_AS","XieH_2016_stool_control","YuJ_2015_stool_CRC","YuJ_2015_stool_control","ZellerG_2014_stool_control","ZellerG_2014_stool_CRC","ZellerG_2014_stool_adenoma")
  AllWMSData_list <- AllWMSData_list[names %in% names_paper]
  save(AllWMSData_list,file = "../data/AllWMSData_list.RData")
} else load("../data/AllWMSData_list.RData")
names(AllWMSData_list)
```

For a total of:

```{r}
cat("HMP16SData \n")
length(HMP16SData_list)

cat("HMPWMSData \n")
length(HMPWMSData_list)

cat("AllWMSData \n")
length(AllWMSData_list)
```

Where *nasal_cavity*, *stool* and *vagina* dataset of AllWMSData_list contain the same samples of *Anterior nares*, *Stool* and *Posterior fornix* of HMPWMSData_list.

# Model estimation

Zero-Inflated Negative Binomial (ZINB), Negative Binomial (NB), Truncated Gaussian hurdle model (MAST), Zero-inflated Gaussian (ZIG), Dirichlet-Multinomial (DM).

(This files could be >200MB, not reported)

```{r, echo=FALSE,message=FALSE,warning=FALSE}
source("additional_functions.R")

if(!file.exists("../data/HMP16SData_models.RData")){
  HMP16SData_models <- lapply(HMP16SData_list,function(ps) fit_models(ps, model.matrix( ~ 1, data = data.frame(ps@sam_data))))
  save(HMP16SData_models,file = "../data/HMP16SData_models.RData")
} else load("../data/HMP16SData_models.RData")

if(!file.exists("../data/HMPWMSData_models.RData")){
  HMPWMSData_models <- lapply(HMPWMSData_list,function(ps) fit_models(ps, model.matrix( ~ 1, data = data.frame(ps@sam_data))))
  save(HMPWMSData_models,file = "../data/HMPWMSData_models.RData")
} else load("../data/HMPWMSData_models.RData")

if(!file.exists("../data/AllWMSData_models.RData")){
  AllWMSData_models <- lapply(AllWMSData_list[1:2],function(ps) fit_models(ps, model.matrix( ~ 1, data = data.frame(ps@sam_data))))
  save(AllWMSData_models,file = "../data/AllWMSData_models.RData")
} else load("../data/AllWMSData_models.RData")
```

## Mean values and Zero Probability

All models are fitted on all datasets. For each dataset, the estimated and the observed count value are compared (Mean Differences), and the estimated probability to have a zero is compared to the real proportion of zeroes (Zero Probability Differences). 

```{r}
HMP16SData_df <- model_to_data.frame(HMP16SData_models)
HMPWMSData_df <- model_to_data.frame(HMPWMSData_models)
AllWMSData_df <- model_to_data.frame(AllWMSData_models)

head(HMP16SData_df)
head(HMPWMSData_df)
head(AllWMSData_df)
```

Goodness of fit is evaluated through Root Mean Square Error (RMSE). 

```{r}
gof_indexes_HMP16S <- compute_RMSE(HMP16SData_df)
gof_indexes_HMPWMS <- compute_RMSE(HMPWMSData_df)
gof_indexes_AllWMS <- compute_RMSE(AllWMSData_df)

head(gof_indexes_HMP16S)
head(gof_indexes_HMPWMS)
head(gof_indexes_AllWMS)
```

# Plot

```{r}
model_colors <- brewer.pal(7,"Set2")
names(model_colors) <- c("ZINB","NB","HURDLE_default","HURDLE_median","ZIG_default","ZIG_median","DM")
```

```{r}
# GOF for all datasets
gof_indexes_list <- list(AllWMS = gof_indexes_AllWMS,
                         HMP16S = gof_indexes_HMP16S,
                         HMPWMS = gof_indexes_HMPWMS)
gof_indexes <- melt(ldply(gof_indexes_list))
gof_indexes$type <- "RMSE"

# Model results for all datasets
df_list <- list(AllWMG = AllWMSData_df,
                HMP16S = HMP16SData_df,
                HMPWMG = HMPWMSData_df)
df_whole <- ldply(lapply(df_list, function(x){
                  estimated <- c(x$EY,x$EY0)
                  observed <- c(x$Y,x$Y0)
                  df_whole <- data.frame(Models = rep(x$Models,2),
                                         Subsites = rep(x$Subsites,2),
                                         Observed = observed,
                                         Estimated = estimated,
                                         Variable = rep(c("MD","ZPD"),each = length(estimated)/2))
}))
```

```{r}
library(ggplot2)

# Methods for plots

MD_ZPD_plot <- function(subsite,variable,dataset){
  df_whole$.id <- factor(df_whole$.id,levels = c("HMP16S","HMPWMG","AllWMG"), labels = c("HMP16S","HMPWMS","AllWMS"))
  ggplot(df_whole[df_whole$Subsites %in% subsite & df_whole$Variable %in% variable & df_whole$.id %in% dataset,],aes(x = Observed, y = Estimated - Observed,color = Models)) + 
    facet_wrap(~ Variable,nrow = 1, scales = "free") + 
    # geom_point(size = 0.001) + 
    geom_smooth(aes(lty = Models),se = FALSE) + 
    # geom_point(size = 1, shape = 21, alpha = 0.3) +
    theme_minimal() +
    theme(legend.position = "none",
          #strip.text = element_blank(),
          axis.text.x = element_text(hjust = 1,angle = 45),
          panel.spacing = unit(0.5, "lines"),
          plot.margin = unit(c(0.1,0.1,0.1,0.1), "cm")) + 
    geom_hline(yintercept = 0, color = "darkgrey") + 
    scale_color_manual(values = model_colors) +
    xlab("Observed values") + 
    ylab("Estimated - Observed") +
    ggtitle(label = paste(variable,"-plot",sep = ""),subtitle = paste(dataset,subsite,sep = "-"))
}

RMSE_plot <- function(variable,type,subsite,dataset){
  ord <- order(gof_indexes[gof_indexes$variable %in% variable & gof_indexes$type %in% type & gof_indexes$Subsites %in% subsite & gof_indexes$.id %in% dataset,"value"])
  ggplot(gof_indexes[gof_indexes$variable %in% variable & gof_indexes$type %in% type & gof_indexes$Subsites %in% subsite & gof_indexes$.id %in% dataset,],aes(x = Models, y = value, fill = Models)) + 
    facet_wrap(variable ~ Subsites, scales = "free",nrow = 1) +
    geom_bar(stat = "identity",position = "dodge") + 
    scale_shape_discrete(solid = FALSE) + 
    scale_x_discrete(limits = gof_indexes[gof_indexes$variable %in% variable & gof_indexes$type %in% type & gof_indexes$Subsites %in% subsite & gof_indexes$.id %in% dataset,"Models"][ord]) +
    theme_minimal() +
    theme(legend.position = "none",
          strip.text = element_blank(),
          axis.text.x = element_text(hjust = 1,angle = 45),
          # axis.text.x = element_blank(),
          # axis.text.y = element_blank(),
          panel.spacing = unit(0, "lines"),
          plot.margin = unit(c(0,0,0,0), "cm")) + 
    geom_hline(yintercept = 0, color = "darkgrey") + 
    labs(shape = "Index") +
    scale_fill_manual(values = model_colors) +
    # coord_flip() +
    xlab("Models") + 
    ylab("RMSE")
}
```

```{r}
g_legend <- get_legend(ggplot(df_whole[df_whole$Subsites %in% "Stool" & df_whole$Variable %in% "MD",],aes(x = Observed, y = Estimated - Observed,color = Models)) + 
                         facet_wrap(~ Variable,nrow = 1, scales = "free") + 
                         # geom_point(size = 0.001) + 
                         geom_smooth(aes(lty = Models),se = FALSE) + 
                         # geom_point(size = 1, shape = 21, alpha = 0.3) +
                         theme_minimal() +
                         theme(legend.position = "bottom") +
                         guides(color = guide_legend(nrow=1,byrow=TRUE)) +
                         scale_color_manual(values = model_colors))

a1 <- MD_ZPD_plot(subsite = "Stool",variable = "MD",dataset = "HMP16S")
a2 <- RMSE_plot(variable = "MD",type = "RMSE",subsite = "Stool",dataset = "HMP16S")
a <- plot_grid(plotlist = list(a1,a2),align = "v",axis = "lr",nrow = 2)

b1 <- MD_ZPD_plot(subsite = "Stool",variable = "MD",dataset = "HMPWMS")
b2 <- RMSE_plot(variable = "MD",type = "RMSE",subsite = "Stool",dataset = "HMPWMS")
b <- plot_grid(plotlist = list(b1,b2),align = "v",axis = "lr",nrow = 2)

d1 <- MD_ZPD_plot(subsite = "Stool",variable = "ZPD",dataset = "HMP16S")
d2 <- RMSE_plot(variable = "ZPD",type = "RMSE",subsite = "Stool",dataset = "HMP16S")
d <- plot_grid(plotlist = list(d1,d2),align = "v",axis = "lr",nrow = 2)

e1 <- MD_ZPD_plot(subsite = "Stool",variable = "ZPD",dataset = "HMPWMS")
e2 <- RMSE_plot(variable = "ZPD",type = "RMSE",subsite = "Stool",dataset = "HMPWMS")
e <- plot_grid(plotlist = list(e1,e2),align = "v",axis = "lr",nrow = 2)
```

```{r, fig.width=10, fig.height=12}
# Heatmap
gof_indexes_ranked <- ddply(gof_indexes,.variables = ~ Subsites + variable + type + .id,.fun = function(x) rank(x[,"value"]))
colnames(gof_indexes_ranked) <- c("Subsites","variable","type","Dataset",as.character(unique(gof_indexes$Models)))
# Average ranks per project name
all_RMSE_summary <- melt(ddply(gof_indexes_ranked,.variables = ~ variable + type + Dataset, function(x) colMeans(x[,5:11])))
colnames(all_RMSE_summary) <- c("Variable","Type","Dataset","Model","Rank")

all_RMSE_summary$Dataset <- factor(all_RMSE_summary$Dataset,levels = c("HMP16S","HMPWMS","AllWMS"),labels = c("HMP16S","HMPWMS","AllWMS"),ordered = TRUE)

# Heatmap values
gof_indexes_values <- ddply(gof_indexes,.variables = ~ Subsites + variable + type + .id,.fun = function(x) x[,"value"])
colnames(gof_indexes_values) <- c("Subsites","variable","type","Dataset",as.character(unique(gof_indexes$Models)))
# Average mean per method
all_RMSE_summary_values <- melt(ddply(gof_indexes_values,.variables = ~ variable + type + Dataset, function(x) colMeans(x[,5:11])))
colnames(all_RMSE_summary_values) <- c("Variable","Type","Dataset","Model","Value")

all_RMSE_summary$Value <- all_RMSE_summary_values$Value


g_legend_heat <- get_legend(ggplot(all_RMSE_summary[all_RMSE_summary$Type == "RMSE" & all_RMSE_summary$Variable == "ZPD",], aes(x = Dataset, y = Model, fill = Rank)) +
  geom_tile(height = 0.8, width = 0.8,) + 
  theme_minimal() + 
  scale_fill_distiller(palette = "RdYlBu") +
  theme(legend.position = "bottom",
        panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_colorbar()) + 
  labs(fill = "Mean Rank:"))

plot_heat <- function(variable,type, digits = 1){
  ord <- order(ddply(all_RMSE_summary[all_RMSE_summary$Type==type & all_RMSE_summary$Variable == variable,],.variables = ~ Model, function(x) mean(x[,"Rank"]))$V1)
  ggplot(all_RMSE_summary[all_RMSE_summary$Type == type & all_RMSE_summary$Variable == variable,], aes(x = Dataset, y = Model, fill = Rank)) +
    geom_tile(height = 0.8, width = 0.8) + 
    geom_text(aes(x = Dataset, y = Model, label=ifelse(Value<0.005,"<0.01",round(Value,digits = digits)))) +
    coord_equal() +
    #facet_wrap(~ Dataset, labeller = label_wrap_gen(width=10)) +
    theme_minimal() + 
    scale_y_discrete(limits = levels(all_RMSE_summary$Model)[ord]) +
    scale_fill_distiller(palette = "RdYlBu") +
    theme(legend.position = "none",
          panel.grid = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle(label = paste("Mean",type,"ranks for",variable),subtitle = "16S and WMS datasets") +
    guides(fill = guide_colorbar()) + 
    labs(fill = "Mean Rank:") + xlab("Dataset")
}

c <- plot_heat(variable = "MD",type = "RMSE", digits = 1)
f <- plot_heat(variable = "ZPD",type = "RMSE", digits = 2)

abde <- plot_grid(plotlist = list(a,b,d,e),nrow = 2,ncol = 2,labels = c("a","b","d","e"))
cf <- plot_grid(plotlist = list(c,f),nrow = 2, labels = c("c","f"))

abde_legend <- plot_grid(plotlist = list(abde,g_legend),nrow = 2, rel_heights = c(1,0.1))
cf_legend <- plot_grid(plotlist = list(cf,g_legend_heat),nrow = 2, rel_heights = c(1,0.1))

fig <- plot_grid(plotlist = list(abde_legend,cf_legend),ncol = 2,rel_widths = c(1,0.6))
fig
```




